Rapid-response Bayesian sky localization for electromagnetic follow-up of gravitational-wave candidates

### 2013-08-02 ms.tex
Small improvements to prose

In the frequency domain, the data recorded by a single \ac{GW} interferometer are
%
\begin{equation}\label{eq:signal-model}
	x_i (\omega) = h_i (\omega; \boldsymbol\theta) + n_i (\omega),
\end{equation}
%
where $h_i (\omega; \boldsymbol\theta)$ is the \ac{GW} signal received by detector $i$, given a parameter vector $\boldsymbol\theta$ that describes the \ac{GW} source, and $n_i (\omega)$ is that detector's \ac{WSS} Gaussian noise with one\nobreakdashes-sided \ac{PSD} $S_i(\omega) = E\left[n_i(\omega)^2\right]$. We shall denote the combined observation from a network of detectors as $\mathbf x (\omega) \equiv \{x_i (\omega)\}_i$.

The likelihood, or the probability of obtaining the observation $\mathbf x$ conditioned on the value of $\boldsymbol\theta$, is the product of Gaussian distributions, one Gaussian for each detector:
%
\begin{equation}\label{eq:gaussian-likelihood}
	\mathcal{L}(\mathbf x; \boldsymbol\theta) = \prod_i p(x_i | \boldsymbol\theta)
		\propto \exp \left[
		- \frac{1}{2} \sum_i \int_0^\infty \frac{\left|x_i (\omega)
			- h_i(\omega; \boldsymbol\theta) \right|^2}{S_i(\omega)} \, \mathrm{d}\omega
	\right].
\end{equation}

A \ac{CBC} source is specified by a vector of extrinsic parameters describing its position and orientation and intrinsic parameters describing the physical properties of the binary components:
%
\begin{equation}
    \boldsymbol\theta = \begin{array}{l@{}l}
            \left[
            \begin{array}{c}
                \alpha \\
                \delta \\
                D_\mathrm{L} \\
                t_\oplus \\
                u \\
                \psi \\
                \phi_c \\
                \hdashline[1pt/1pt]
                m_1 \\
                m_2 \\
                \mathbf S_1 \\
                \mathbf S_2
            \end{array}
            \right] &
            \begin{array}{l}
                \left.
                \begin{array}{p{4cm}}
                    right ascension \\
                    declination \\
                    luminosity distance \\
                    arrival time at geocenter \\
                    $\cos(\textrm{inclination angle})$ \\
                    polarization angle \\
                    coalescence phase \\
                    \hdashline[1pt/1pt]
                \end{array}
                \quad \right\} \textnormal{extrinsic parameters}, \boldsymbol\theta_\mathrm{ex} \\
                \left.
                \begin{array}{p{4cm}}
                    first component's mass \\
                    second component's mass \\
                    first component's spin \\
                    second component's spin
                \end{array}
                \quad \right\} \textnormal{intrinsic parameters}, \boldsymbol\theta_\mathrm{in}
            \end{array}
        \end{array}
\end{equation}
%
Even this list of parameters involves some simplifying assumptions. First, for \ac{BNS} systems, we omit binary eccentricity as an intrinsic parameter because tidal interaction of the progenitor stars~\cite{0004-637X-572-1-407} and later \ac{GW} emission~\cite{PhysRev.136.B1224} circularize the orbit long before the inspiral enters \ac{LIGO}'s frequency range of $\sim$10\nobreakdashes--1000~kHz. Eccentricity would play a major role in the evolution and waveforms of \ac{NSBH} and \ac{BBH} sources formed by dynamical capture~\cite{PhysRevD.87.043004}. Tidal deformabilities of the \acp{NS} are omitted because the effect of tides on the inspiral would only be detectable by an Einstein Telescope-class \ac{GW} observatory~\cite{PhysRevD.81.123016}. Furthermore, in \ac{GW} detection efforts, especially those focused on \ac{BNS} systems, the component spins $\mathbf{S}_1$ and $\mathbf{S}_2$ are often assumed to be nonprecessing and aligned with the system's total angular momentum and condensed to a single scalar parameter $\chi$, or even neglected entirely, $\mathbf{S}_1 = \mathbf{S}_2 = 0$. In this work we will not consider spin precession.

@@

The prevailing technique for detection of \acp{GW} from \acp{CBC} is to realize a \ac{MLE} from \Eref{eq:gaussian-likelihood} with the above signal model. This takes the form of a bank of matched filters, or template waveforms that are continually convolved with the incoming data stream. The matched filter or \ac{ML} estimates are given by
%
\begin{eqnarray}
\hat{\boldsymbol\theta}_i, \{\hat\tau_i\}_i &=& \argmax_{\boldsymbol\theta_\mathrm{in}, \{\hat\tau_i\}_i} \sum_i \frac{1}{{\sigma_i}^2} \left| \int_0^\infty \frac{h^*(\omega; \boldsymbol\theta_\mathrm{in}) x_i(\omega) e^{-i \omega \tau_i}}{S_i(\omega)} \,\mathrm{d}\omega \right|^2 \\
\hat\rho_i &=& \frac{1}{\sigma_i} \left| \int_0^\infty \frac{h^*(\omega; \hat{\boldsymbol\theta}_\mathrm{in}) x_i(\omega) e^{-i \omega \hat\tau_i}}{S_i(\omega)} \,\mathrm{d}\omega \right| \\
\hat\gamma_i &=& \arg \int_0^\infty \frac{h^*(\omega; \hat{\boldsymbol\theta}_\mathrm{in}) x_i(\omega) e^{-i \omega \hat\tau_i}}{S_i(\omega)} \,\mathrm{d}\omega,
\end{eqnarray}
%
where the $^*$ denotes a complex conjugate.

@@

Computing the marginal posterior almost always involves many-dimensional and often ill-behaved integrals.\marginpar{More about \ac{MCMC} methods here.}

@@

In order to estimate the \ac{ML} estimator accuracy for a matched filter, we will momentarily consider the likelihood for a single detector, and suppress the intrinsic degrees of freedom:
%
\begin{equation}\label{eq:gaussian-likelihood}
	\fl\mathcal{L}(x_i; \rho_i, \gamma_i, \tau_i) = p(x_i | \rho_i, \gamma_i, \tau_i)
		\propto \exp \left[
		- \frac{1}{2} \int_0^\infty \frac{\left|x_i (\omega)
			- h_i(\omega; \rho_i, \gamma_i, \tau_i) \right|^2}{S_i(\omega)} \, \mathrm{d}\omega
	\right],
\end{equation}
%
with $h_i(\omega; \rho_i, \gamma_i, \tau_i)$ given by \Eref{eq:signal-model}.

@@

The likelihood in \Eref{eq:gaussian-likelihood} is the quantity that is maximized by the matched\nobreakdashes-filter estimator. However, just because we have the covariance matrix defined by \Eref{eq:covariance-matrix} does not mean that the distribution of estimator errors is multivariate Gaussian. Nonetheless, it is straightforward to qualitatively describe the distribution.

>>>>

In the frequency domain, the strain observed by a single \ac{GW} interferometer is
%
\begin{equation}\label{eq:signal-model}
	x_i (\omega) = h_i (\omega; \boldsymbol\theta) + n_i (\omega),
\end{equation}
%
where $h_i (\omega; \boldsymbol\theta)$ is the \ac{GW} signal given a parameter vector $\boldsymbol\theta$ that describes the \ac{GW} source, and $n_i (\omega)$ is that detector's \ac{WSS} Gaussian noise with one\nobreakdashes-sided \ac{PSD} $S_i(\omega) = E\left[n_i(\omega)^2\right]$. We shall denote the combined observation from a network of detectors as $\mathbf x (\omega) \equiv \{x_i (\omega)\}_i$.

The likelihood, or the probability of obtaining the observation $\mathbf x$ conditioned on the value of $\boldsymbol\theta$, is a product of Gaussian distributions:
%
\begin{equation}\label{eq:gaussian-likelihood}
	\mathcal{L}(\mathbf x; \boldsymbol\theta) = \prod_i p(x_i | \boldsymbol\theta)
		\propto \exp \left[
		- \frac{1}{2} \sum_i \int_0^\infty \frac{\left|x_i (\omega)
			- h_i(\omega; \boldsymbol\theta) \right|^2}{S_i(\omega)} \, \mathrm{d}\omega
	\right].
\end{equation}

A \ac{CBC} source is specified by a vector of extrinsic parameters describing its position and orientation and intrinsic parameters describing the physical properties of the binary components:
%
\begin{equation}
    \boldsymbol\theta = \begin{array}{l@{}l}
            \left[
            \begin{array}{c}
                \alpha \\
                \delta \\
                D_\mathrm{L} \\
                t_\oplus \\
                u \\
                \psi \\
                \phi_c \\
                \hdashline[1pt/1pt]
                m_1 \\
                m_2 \\
                \mathbf S_1 \\
                \mathbf S_2
            \end{array}
            \right] &
            \begin{array}{l}
                \left.
                \begin{array}{p{4cm}}
                    right ascension \\
                    declination \\
                    luminosity distance \\
                    arrival time at geocenter \\
                    $\cos(\textrm{inclination angle})$ \\
                    polarization angle \\
                    coalescence phase \\
                    \hdashline[1pt/1pt]
                \end{array}
                \quad \right\} \textnormal{extrinsic parameters}, \boldsymbol\theta_\mathrm{ex} \\
                \left.
                \begin{array}{p{4cm}}
                    first component's mass \\
                    second component's mass \\
                    first component's spin \\
                    second component's spin
                \end{array}
                \quad \right\} \textnormal{intrinsic parameters}, \boldsymbol\theta_\mathrm{in}
            \end{array}
        \end{array}
\end{equation}
%
This list of parameters involves some simplifying assumptions. Eccentricity is omitted: although it does play a major role in the evolution and waveforms of \ac{NSBH} and \ac{BBH} sources formed by dynamical capture~\cite{PhysRevD.87.043004}, \ac{BNS} systems formed by binary stellar evolution should circularize due to tidal interaction~\cite{0004-637X-572-1-407} and later \ac{GW} emission~\cite{PhysRev.136.B1224} long before the inspiral enters \ac{LIGO}'s frequency range of $\sim$10\nobreakdashes--1000~kHz. Tidal deformabilities of the \acp{NS} are omitted because the signal imprinted by the companions' material properties is so small that it will only be detectable by an Einstein Telescope\nobreakdashes-class \ac{GW} observatory~\cite{PhysRevD.81.123016}. Furthermore, in \ac{GW} detection efforts, especially those focused on \ac{BNS} systems, the component spins $\mathbf{S}_1$ and $\mathbf{S}_2$ are often assumed to be nonprecessing and aligned with the system's total angular momentum and condensed to a single scalar parameter $\chi$, or even neglected entirely, $\mathbf{S}_1 = \mathbf{S}_2 = 0$.

@@

The prevailing technique for detection of \acp{GW} from \acp{CBC} is to realize a \ac{MLE} from the likelihood in \Eref{eq:gaussian-likelihood} and the signal model in \Eref{eq:signal-model}. Concretely, this results in a bank of matched filters, or template waveforms that are convolved with the incoming data stream. The matched filter or \ac{ML} parameter estimates are given by
%
\begin{eqnarray}
\hat{\boldsymbol\theta}_i, \{\hat\tau_i\}_i &=& \argmax_{\boldsymbol\theta_\mathrm{in}, \{\hat\tau_i\}_i} \sum_i \frac{1}{{\sigma_i}^2} \left| \int_0^\infty \frac{h^*(\omega; \boldsymbol\theta_\mathrm{in}) x_i(\omega) e^{-i \omega \tau_i}}{S_i(\omega)} \,\mathrm{d}\omega \right|^2 \\
\hat\rho_i &=& \frac{1}{\sigma_i} \left| \int_0^\infty \frac{h^*(\omega; \hat{\boldsymbol\theta}_\mathrm{in}) x_i(\omega) e^{-i \omega \hat\tau_i}}{S_i(\omega)} \,\mathrm{d}\omega \right| \\
\hat\gamma_i &=& \arg \int_0^\infty \frac{h^*(\omega; \hat{\boldsymbol\theta}_\mathrm{in}) x_i(\omega) e^{-i \omega \hat\tau_i}}{S_i(\omega)} \,\mathrm{d}\omega,
\end{eqnarray}
%
where the $^*$ denotes a complex conjugate.

@@

Bayesian parameter estimation has many advantages, including broad generality and the ability to make probabilistically meaningful statements even with very low \ac{SNR} measurements. However, in problems of even modest complexity, the marginalization step burdens the analysis with nasty, many-dimensional, ill-behaved integrals. Almost all modern parameter estimation tactics are variations of \ac{MCMC} integration. Though powerful, \ac{MCMC} is slow and non-deterministic. With the most sophisticated \ac{CBC} parameter estimation codes, it still takes days to process a single event.

@@

In order to estimate the \ac{ML} estimator accuracy for a matched filter, we will momentarily consider the likelihood for a single detector, and suppress the intrinsic degrees of freedom:
%
\begin{equation}\label{eq:gaussian-likelihood-spa}
	\fl\mathcal{L}(x_i; \rho_i, \gamma_i, \tau_i) = p(x_i | \rho_i, \gamma_i, \tau_i)
		\propto \exp \left[
		- \frac{1}{2} \int_0^\infty \frac{\left|x_i (\omega)
			- h_i(\omega; \rho_i, \gamma_i, \tau_i) \right|^2}{S_i(\omega)} \, \mathrm{d}\omega
	\right],
\end{equation}
%
with $h_i(\omega; \rho_i, \gamma_i, \tau_i)$ given by \Eref{eq:signal-model}.

@@

The likelihood in \Eref{eq:gaussian-likelihood-spa} is the quantity that is maximized by the matched\nobreakdashes-filter estimator. However, just because we have the covariance matrix defined by \Eref{eq:covariance-matrix} does not mean that the distribution of estimator errors is multivariate Gaussian. Nonetheless, it is straightforward to qualitatively describe the distribution.
